---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
- Load necessary packages
- if csv file does not exist unzip activity.zip
```{r echo=TRUE}
suppressMessages(library(tidyverse))
library(lattice)
if(!file.exists("activity.csv")){unzip("activity.zip")}
activity = read.csv("activity.csv")
head(activity)
```


## What is mean total number of steps taken per day?
```{r}
daily_activity = activity[!is.na(activity),] %>%
  group_by(date) %>%
  summarise(daily_steps=sum(steps))
```
- Mean number of steps: `r round(mean(daily_activity$daily_steps, na.rm=TRUE), 2)`
- Median number of steps: `r round(median(daily_activity$daily_steps, na.rm=TRUE), 2)`

```{r echo=TRUE}
hist(daily_activity$daily_steps, main="Histogram: Total Number of Steps per day")
mean(daily_activity$daily_steps, na.rm=TRUE)
median(daily_activity$daily_steps, na.rm=TRUE)
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r echo=TRUE}
summary_activity = activity[!is.na(activity$steps), ] %>%
  group_by(interval) %>%
  summarise(AvgSteps = mean(steps))

max_step_interval <- summary_activity[summary_activity$AvgSteps == max(summary_activity$AvgSteps), "interval" ][[1]]

with(summary_activity, 
     plot(x= interval,
          y=AvgSteps,
          type="l",
          main="Average Daily Steps by Interval" )
)
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r echo=TRUE}

max_step_interval
```


## Imputing missing values
1. Calc number of NA's
```{r echo=TRUE}
sum(is.na(activity$steps))
```

2. Fill NA's with mean from interval
```{r echo=TRUE}
na_fill_mean = function(interval){
  result = seq_len(length(interval))
  for(i in 1:length(interval)){
    result[i] = mean(activity$steps[activity$interval==interval[i]], na.rm=TRUE)
      }
  return(as.vector(result))
}

```
3. Create copy of dataset replacing NAs
```{r echo=TRUE}
clean_activity <- activity
clean_activity$steps[is.na(activity$steps)] = na_fill_mean(activity[is.na(activity),"interval"])
str(clean_activity)
str(activity)
```
4. Historgram, mean and median of clean dataset
```{r}
daily_clean_activity = clean_activity[!is.na(activity),] %>%
  group_by(date) %>%
  summarise(steps=sum(steps))
```

```{r echo=TRUE}
hist(daily_clean_activity$steps, main="Histogram: Total Number of Steps per day")
mean(daily_clean_activity$steps, na.rm=TRUE)
median(daily_clean_activity$steps, na.rm=TRUE)

```
- change in mean: `r mean(daily_clean_activity$steps, na.rm=TRUE) - mean(daily_activity$daily_steps, na.rm=TRUE)`
- change in median: `r median(daily_clean_activity$steps, na.rm=TRUE) - median(daily_activity$daily_steps, na.rm=TRUE)`

## Are there differences in activity patterns between weekdays and weekends?

1. Create factor variable for weekday vs weekend

```{r} 
WEEKEND = c("Saturday", "Sunday")
clean_activity$weekend = weekdays(as.Date(clean_activity$date)) %in% WEEKEND
clean_activity$weekend[clean_activity$weekend==TRUE] = "weekend"
clean_activity$weekend[clean_activity$weekend==FALSE] = "weekday"
clean_activity$weekend = as.factor(clean_activity$weekend)
str(clean_activity)
```

2. Plot time series comparing weekday vs weekend

```{r echo=TRUE}
summary_clean_activity = clean_activity %>%
  group_by(interval, weekend) %>%
  summarise(AvgSteps = mean(steps))

xyplot(AvgSteps ~ interval | weekend,
       data = summary_clean_activity,
       type="l",
       layout = c(1,2),
       main="Average Daily Steps by Interval" 
       )

```
